Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations

### 2023-09-07 index.tex

\title{Graph embedding and transfer learning can help predict potential
species interaction networks despite data limitations}
\author{Tanya Strydom \and Salomé Bouskila \and Francis
Banville \and Ceres Barros \and Dominique Caron \and Maxwell J.
Farrell \and Marie-Josée Fortin \and Victoria Hemming \and Benjamin
Mercier \and Laura J. Pollock \and Rogini Runghen \and Giulio V. Dalla
Riva \and Timothée Poisot}
\date{}

\begin{document}
\maketitle
The ability to infer \emph{potential} interactions could serve as a
significant breakthrough in our ability to conceptualize species
interaction networks over large spatial scales (Hortal et al. 2015).
Reliable inferences would not only boost our understanding of the
structure of species interaction networks, but also increase the amount
of information that can be used for biodiversity management. In a recent
overview of the field of ecological network prediction, Strydom et al.
(2021) identified two challenges of interest to the prediction of
interactions at large scales. First, there is a relative scarcity of
relevant data in most places globally -- which, due to the limitations
in most predictive methods, restricts the ability to infer interactions
to locations where it is least required (\emph{i.e.} regions where we
already have interaction data) leaving us unable to make inference in
data scarce regions (where we most need it); second, accurate predictors
are important for accurate predictions, and the lack of methods that can
leverage a small amount of \emph{accurate} data is a serious impediment
to our predictive ability. In this contribution, we highlight the power
of viewing (and constructing) metawebs as \emph{probabilistic} objects
in the context of low-probability interactions, discuss how a family of
machine learning tools (graph embeddings and transfer learning) can be
used to overcome data limitations to metaweb inference, and highlight
how the use of metawebs introduces important questions for the field of
network ecology.

In most places, our most reliable biodiversity knowledge is that of a
species pool where a set of potentially interacting species in a given
area could occur: through the analysis of databases like the Global
Biodiversity Information Facility (GBIF) or the International Union for
the Conservation of Nature (IUCN), it is possible to construct a list of
species for a region of interest. Following the definition of Dunne
(2006), a metaweb is the ecological network analogue to the species
pool; specifically, it inventories all \emph{potential} interactions
between species for a spatially delimited area (and so captures the
\(\gamma\) diversity of interactions as per Poisot et al. (2012)).
However, inferring the potential interactions between these species
still remains a challenge. And yet, because the metaweb represents the
joint effect of functional, phylogenetic, and macroecological processes
(Morales-Castilla et al. 2015; Carlson et al. 2022; Morales-Castilla et
al. 2021), it holds valuable ecological information. Specifically, it
represents the ``upper bounds'' on what the composition of the local
networks, given a local species pool, can be (see \emph{e.g.} McLeod et
al. 2021); this information can help evaluate the ability of ecological
assemblages to withstand the effects of, for example, climate change
(Fricke et al. 2022). These local networks may be reconstructed given an
appropriate knowledge of local species composition and provide
information on the structure of networks at finer spatial scales. This
has been done for example for tree-galler-parasitoid systems (Gravel et
al. 2018), fish trophic interactions (Albouy et al. 2019), tetrapod
trophic interactions (J. Braga et al. 2019; O'Connor et al. 2020), and
crop-pest networks (Grünig et al. 2020).

The metaweb itself is not a prediction of local networks at specific
locations within the spatial area it covers: it will have a different
structure, notably by having a larger connectance (see \emph{e.g.} Wood
et al. 2015) and complexity (see \emph{e.g.} Galiana et al. 2022), than
any of these local networks. These local networks (which capture the
\(\alpha\) diversity of interactions) are a subset of the metaweb's
species and its realized interactions, and have been called ``metaweb
realizations'' (Poisot, Stouffer, and Gravel 2015). Differences between
local networks and their metawebs are due to chance, species abundance
and co-occurrence, local environmental conditions, and local
distribution of functional traits, among others. Specifically, although
co-occurrence can be driven by interactions (Cazelles et al. 2016),
co-occurrence alone is not a predictor of interactions (Blanchet,
Cazelles, and Gravel 2020; Thurman et al. 2019), and therefore the lack
of co-occurrence cannot be used to infer the lack of a feasible
interaction. Yet, recent results by Saravia et al. (2021) strongly
suggested that local (metaweb) realizations only respond weakly to local
conditions: instead, they reflect constraints inherited by the structure
of their metaweb. This sets up the core goal of predictive network
ecology as the prediction of metaweb structure, as it is required to
accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas, Park, and
Drake (2017) suggested that most interactions (links) in ecological
networks are cryptic, \emph{i.e.} uncommon or hard to observe. This
argument echoes Jordano (2016): sampling ecological interactions is
difficult because it requires first the joint observation of two
species, and then the observation of their interaction. In addition, it
is generally expected that weak or rare interactions will be more
prevalent in networks than common or strong interactions (Csermely
2004), compared to strong, persistent interactions; this is notably the
case in food chains, wherein many weaker interactions are key to the
stability of a system (Neutel, Heesterbeek, and de Ruiter 2002). In the
light of these observations, we expect to see an over-representation of
low-probability (hereafter rare) interactions under a model that
accurately predicts interaction probabilities.

Yet, the original metaweb definition, and indeed most past uses of
metawebs, was based on the presence/absence of interactions. Moving
towards \emph{probabilistic} metawebs, by representing interactions as
Bernoulli events (see \emph{e.g.} Poisot et al. 2016), offers the
opportunity to weigh these rare interactions appropriately. The inherent
plasticity of interactions is important to capture: there have been
documented instances of food webs undergoing rapid collapse/recovery
cycles over short periods of time (\emph{e.g.} Pedersen et al. 2017).
Furthermore, because the structure of the metaweb cannot be known in
advance, it is important to rely on predictive tools that do not assume
a specific network topology for link prediction (Gaucher, Klopp, and
Robin 2021), but are able to work on generalizations of the network.
These considerations emphasize why metaweb predictions should focus on
quantitative (preferentially probabilistic) predictions, and this should
constrain the suite of models that are appropriate for prediction.
Binary classifiers based on probabilities have an extremely robust
methodology to validate them, and this applies naturally to the
prediction of interactions (Poisot 2023).

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al. 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between (Catchen et al. 2023). Although metawebs can
(and in practice likely do) include false positives, these are
statistically negligible compared to the false negatives. Furthermore,
Strydom et al. (2022) shows that t-SVD embedding is extremely robust to
(and able to detect) the presence of false positives. In a sense, that
most ecological interactions are elusive can call for a slightly
different approach to sampling: once the common interactions are
documented, the effort required in documenting each rare interaction
will increase exponentially (Jordano 2016). Recent proposals in other
fields relying on machine learning approaches emphasize the idea that
algorithms meant to predict, through the assumption that they
approximate the process generating the data, can also act as data
generators (Hoffmann et al. 2019). High quality observational data can
be used to infer core rules underpinning network structure, and be
supplemented with synthetic data coming from predictive models trained
on them, thereby increasing the volume of information available for
analysis. Indeed, Strydom et al. (2021) suggested that knowing the
metaweb may render the prediction of local networks easier, because it
fixes an ``upper bound'' on which interactions can exist. In this
context, a probabilistic metaweb represents an aggregation of
informative priors on the biological feasibility of interactions, which
is usually hard to obtain yet has possibly the most potential to boost
our predictive ability of local networks (Bartomeus 2013; Bartomeus et
al. 2016). This would represent a departure from simple rules expressed
at the network scale (\emph{e.g.} Williams and Martinez 2000) to a view
of network prediction based on learning the rules that underpin
interactions \emph{and} their variability (Anubhav Gupta, Furrer, and
Petchey 2022).

@@

\caption{\label{fig-embedding}The embedding process (\textbf{A}) can
help to identify links (interactions) that may have been missed within
the original community (represented by the orange dashed arrows,
\textbf{B}). Transfer learning (\textbf{D}) allows for the prediction
links (interactions) even when novel species (\textbf{C}) are included
alongside the original community. This is achieved by learning using
other relevant predictors (\emph{e.g.} traits) in conjunction with the
known interactions to infer latent values (\textbf{E}). Ultimately this
allows us to predict links (interactions) for species external from the
original sample (blue dashed arrows) as well as missing within sample
links (\textbf{F}). Within this context the predicted (and original)
networks as well as the ecological predictors used (green boxes) are
products that can be quantified through measurements in the field,
whereas the embedded as well as imputed matrices (purple box) are
representative of a decomposition of the interaction matrices onto the
embedding space}

@@

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Box 1 - Graph Neural Networks}, colback=white, rightrule=.15mm, coltitle=black, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0, breakable, bottomtitle=1mm, titlerule=0mm, toptitle=1mm, arc=.35mm, toprule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]

@@

different methods on a problem-specific basis, the fact that a lot of

@@

general method of SVD. Note further that tSNE has been included because

@@

tSNE & nodes & statistical divergence & Hinton and Roweis (2002) &

@@

In Figure~\ref{fig-illustration-1}, we focus on some statistical checks
of the embedding. In panel \textbf{A}, we show that the averaged \(L_2\)
loss (\emph{i.e.} the mean of squared errors) between the empirical and
reconstructed metaweb decreases when the number of dimensions (rank) of
the subspace increases, with an inflection at 39 dimensions (out of 120
initially) according to the finite differences method. As discussed by
Runghen, Stouffer, and Dalla Riva (2021), there is often a trade-off
between the number of dimensions to use (more dimensions are more
computationally demanding) and the quality of the representation. In
panel \textbf{B}, we show the increase in cumulative variance explained
at each rank, and visualize that using 39 ranks explains about 70\% of
the variance in the empirical metaweb. This is a different information
from the \(L_2\) loss (which is averaged across interactions), as it
works on the eigenvalues of the embedding, and therefore captures
higher-level features of the network. In panel \textbf{C}, we show
positions of hosts and parasites on the first two dimensions of the left
and right subspaces. Note that these values largely skew negative,
because the first dimensions capture the coarse structure of the
network: most pairs of species do not interact, and therefore have
negative values. Finally in panel \textbf{D}, we show the predicted
weight (\emph{i.e.} the result of the multiplication of the RDGP
subspaces at a rank of 39) as a function of whether the interactions are
observed, not-observed, or unknown due to lack of co-occurrence in the
original dataset. This reveals that the observed interactions have
higher predicted weights, although there is some overlap; the usual
approach to identify potential interactions based on this information
would be a thresholding analysis, which is outside the scope of this
manuscript (and is done in the papers cited in this illustration).
Because the values returned from RDPG are not bound to the unit
interval, we performed a clamping of the weights to the unit space,
showing a one-inflation in documented interactions, and a zero-inflation
in other species pairs. This last figure shows that species pairs with
no documented co-occurrence have weights that are not distinguishable
from species pairs with no documented interactions; in other words,
looking at the embedding, species that do not co-occur are not easily
distinguished from species that do not interact. This suggests that (as
befits a host-parasite model) the ability to interact is a strong
predictor of co-occurrence.

@@

The results of Figure~\ref{fig-illustration-1} show that we can extract
an embedding of the metaweb that captures enough variance to be
relevant; specifically, this is true for both \(L_2\) loss (indicating
that RDPG is able to capture pairwise processes) and the cumulative
variance explained (indicating that RDPG is able to capture
network-level structure). Therefore, in Figure~\ref{fig-illustration-2},
we relate the values of latent variables for hosts to different
ecologically-relevant data. In panel \textbf{A}, we show that host with
a higher value on the first dimension have fewer parasites. This relates
to the body size of hosts in the \emph{PanTHERIA} database (Jones et al.
2009), as shown in panel \textbf{B}: interestingly, the position on the
first axis is only weakly correlated to body mass of the host; this
matches well established results showing that body size/mass is not
always a direct predictor of parasite richness in terrestrial mammals
(Morand and Poulin 1998), a result we observe in panel \textbf{C}.
Finally, in panel \textbf{D}, we can see how different taxonomic
families occupy different positions on the first axis, with \emph{e.g.}
Sciuridae being biased towards higher values. These results show how we
can look for ecological informations in the output of network
embeddings, which can further be refined into the selection of
predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions
between species at a large spatial scale, typically a scale large enough
to be considered of biogeographic relevance (indeed, many of the
examples covered in the introduction span areas larger than a country,
some of them global). But as Herbert (1965) rightfully pointed out,
``{[}y{]}ou can't draw neat lines around planet-wide problems''; any
inference of a metaweb must therefore contend with several novel,
interwoven, families of problems. In this section, we outline three that
we think are particularly important, and can discuss how they may
addressed with subsequent data analysis or simulations, and how they
emerge in the specific context of using embeddings; some of these issues
are related to the application of these methods at the science-policy
interface. Adressing these considerations as part of the methodological
discussion is particularly important, as the construction of metawebs
can perpetuate legacies of biases in data (Box 2).

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic
point of view, the chances of finding another area with enough related
species (through phylogenetic relatedness or similarity of functional
traits) to make a reliable inference decreases. This is because transfer
requires similarity (Figure~\ref{fig-embedding}). A diagnostic for the
lack of similar species would likely be large confidence intervals
during estimation of the values in the low-rank space. In other words,
the representation of the original graph is difficult to transfer to the
new problem. Alternatively, if the initial metaweb is too large
(taxonomically), then the resulting embeddings would need to represent
interactions between taxonomic groups that are not present in the new
location. This would lead to a much higher variance in the starting
dataset, and to under-dispersion in the target dataset, resulting in the
potential under or over estimation of the strength of new predicted
interactions. Llewelyn et al. (2022) provided compelling evidence for
these situations by showing that, even at small spatial scales, the
transfer of information about interactions becomes more challenging when
areas rich with endemic species are considered. The lack of well
documented metawebs is currently preventing the development of more
concrete guidelines. The question of phylogenetic relatedness and
distribution is notably relevant if the metaweb is assembled in an area
with mostly endemic species (\emph{e.g.} a system that has undergone
recent radiation or that has remained in isolation for a long period of
time might not have an analogous system with which to draw knowledge
from), and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to distribution of species in the geographic or phylogenetic
space, it can certainly be approached through assessing the performance
of embedding transfer in simulated starting/target species pools.

@@

Predictive approaches in ecology, regardless of the scale at which they
are deployed and the intent of their deployment, originate in the
framework that contributed to the ongoing biodiversity crisis (Adam
2014) and reinforced environmental injustice (Choudry 2013; Domínguez
and Luoma 2020). The risk of embedding this legacy in our models is
real, especially when the impact of this legacy on species pools is
being increasingly documented. This problem can be addressed by
re-framing the way we interact with models, especially when models are
intended to support conservation actions. Particularly on territories
that were traditionally stewarded by Indigenous people, we must
interrogate how predictive approaches and the biases that underpin them
can be put to task in accompanying Indigenous principles of land
management (Eichhorn, Baker, and Griffiths 2019; No'kmaq et al. 2021).
The discussion of ``algorithm-in-the-loop'' approaches that is now
pervasive in the machine learning community provides examples of why
this is important. Human-algorithm interactions are notoriously
difficult and can yield adverse effects (Green and Chen 2019; Stevenson
and Doleac 2021), suggesting the need to systematically study them for
the specific purpose of, here, biodiversity governance. Improving the
algorithmic literacy of decision makers is part of the solution
(\emph{e.g.} Lamba et al. 2019; Mosebo Fernandes et al. 2020), as we can
reasonably expect that model outputs will be increasingly used to drive
policy decisions (Weiskopf et al. 2022). Our discussion of these
approaches need to go beyond the technical and statistical, and into the
governance consequences they can have. To embed data also embeds
historical and contemporary biases that acted on these data, both
because they shaped the ecological processes generating them, and the
global processes leading to their measurement and publication. For a
domain as vast as species interaction networks, these biases exist at
multiple scales along the way, and a challenge for prediction is not
only to develop (or adopt) new quantitative tools, but to assess the
behavior of these tools in the proper context.

@@

Although promising, the application of embeddings to metaweb prediction
will require to address several challenges. First, there is a need to
understand how to define a metaweb as a single, cohesive, unit of
ecological organisation. This is likely to admit very different answers
based on the specific taxonomic group, temporal and spatial resolution,
and question being investigated. Second, there is a need to understand
the scale at which these predictions are relevant. Although we have
documented many cases of using embedding to fill gaps in the metaweb,
these techniques can likely be brought into a spatial (and possibly
temporal) context. The validation of these predictions will have to
proceed jointly with empirical sampling of interactions, but also with
the design of downsampling methods. Finally, there is a need for a
greater understanding of how biases in the data propagate to the
predictions. Because the volume of metawebs is currently low, and
because graph embeddings have not been commonly applied, we anticipate
that this discussion will take place organically in the coming years.

>>>>

\title{Graph embedding and transfer learning can help predict potential
species interaction networks despite data limitations}
\author{Tanya Strydom \and Salomé Bouskila \and Francis
Banville \and Ceres Barros \and Dominique Caron \and Maxwell J.
Farrell \and Marie-Josée Fortin \and Benjamin Mercier \and Laura J.
Pollock \and Rogini Runghen \and Giulio V. Dalla Riva \and Timothée
Poisot}
\date{}

\begin{document}
\maketitle
\begin{abstract}
\begin{itemize}
\tightlist
\item
  Metawebs (networks of potential interactions within a species pool)
  are a powerful abstraction to understand how large-scale species
  interaction networks are structured.
\item
  Because metawebs are typically expressed at large spatial and
  taxonomic scales, assembling them is a tedious and costly process;
  predictive methods can help circumvent the limitations in data
  deficiencies, by providing a first approximation of metawebs.''
\item
  One way to improve our ability to predict metawebs is to maximize
  available information by using graph embeddings, as opposed to an
  exhaustive list of species interactions. Graph embedding is an
  emerging field in machine learning that holds great potential for
  ecological problems.
\item
  Here, we outline how the challenges associated with inferring metawebs
  line-up with the advantages of graph embeddings; followed by a
  discussion as to how the choice of the species pool has consequences
  on the reconstructed network, specifically as to the role of
  human-made (or arbitrarily assigned) boundaries and how these may
  influence ecological hypotheses.
\end{itemize}
\end{abstract}
The ability to infer potential biotic interactions could serve as a
significant breakthrough in our ability to conceptualize networks over
large spatial scales (Hortal et al. 2015). Reliable inferences would not
only boost our understanding of the structure of species interaction
networks, but also increase the amount of information that can be used
for biodiversity management. In a recent overview of the field of
ecological network prediction, Strydom et al. (2021) identified two
challenges of interest to the prediction of interactions at large
scales. First, there is a relative scarcity of relevant data in most
places globally -- which, due to the limitations in most predictive
methods, restricts the ability to infer interactions to locations where
it is least required (\emph{i.e.} regions where we already have
interaction data) leaving us unable to make inference in data scarce
regions (where we most need it); second, accurate predictors are
important for accurate predictions, and the lack of methods that can
leverage a small amount of \emph{accurate} data is a serious impediment
to our predictive ability. In this contribution, we (i) highlight the
power of viewing (and constructing) metawebs as \emph{probabilistic}
objects in the context of low-probability interactions, (ii) discuss how
a family of machine learning tools (graph embeddings and transfer
learning) can be used to overcome data limitations to metaweb inference,
and (iii) highlight how the use of metawebs introduces important
questions for the field of network ecology.

In most places, our most reliable biodiversity knowledge is that of a
species pool where a set of potentially interacting species in a given
area could occur: through the analysis of databases like the Global
Biodiversity Information Facility (GBIF) or the International Union for
the Conservation of Nature (IUCN), it is possible to construct a list of
species for a region of interest. Following the definition of Dunne
(2006), a metaweb is the ecological network analogue to the species
pool; specifically, it inventories all \emph{potential} interactions
between species for a spatially delimited area (and so captures the
\(\gamma\) diversity of interactions as per Poisot et al. (2012)).
However, inferring the potential interactions between these species
still remains a challenge. And yet, the metaweb holds valuable
ecological information: it represents the joint effect of functional,
phylogenetic, and macroecological processes (Morales-Castilla et al.
2015; Carlson et al. 2022; Morales-Castilla et al. 2021). Specifically,
it represents the ``upper bounds'' on what the composition of the local
networks, given a local species pool, can be (see \emph{e.g.} McLeod et
al. 2021); this information can help evaluate the ability of ecological
assemblages to withstand the effects of, for example, climate change
(Fricke et al. 2022). These local networks may be reconstructed given an
appropriate knowledge of local species composition and provide
information on the structure of networks at finer spatial scales. This
has been done for example for tree-galler-parasitoid systems (Gravel et
al. 2018), fish trophic interactions (Albouy et al. 2019), terrestrial
tetrapod trophic interactions (J. Braga et al. 2019; O'Connor et al.
2020), and crop-pest networks (Grünig et al. 2020).

The metaweb itself is not a prediction of local networks at specific
locations within the spatial area it covers: it will have a different
structure, notably by having a larger connectance (see \emph{e.g.} Wood
et al. 2015) and complexity (see \emph{e.g.} Galiana et al. 2022), than
any of these local networks. Local networks (which capture the
\(\alpha\) diversity of interactions) are a subset of the metaweb's
species and its realized interactions, and have been called ``metaweb
realizations'' (Poisot, Stouffer, and Gravel 2015). Differences between
local networks and their metawebs are due to chance, species abundance
and co-occurrence, local environmental conditions, and local
distribution of functional traits, among others. Specifically, although
co-occurrence can be driven by interactions (Cazelles et al. 2016),
co-occurrence alone is not a predictor of interactions (Blanchet,
Cazelles, and Gravel 2020; Thurman et al. 2019), and therefore the lack
of co-occurrence cannot be used to infer the lack of a feasible
interaction. Yet, recent results by Saravia et al. (2021) strongly
suggested that local (metaweb) realizations only respond weakly to local
conditions: instead, they reflect constraints inherited by the structure
of their metaweb. This sets up the core goal of predictive network
ecology as the prediction of metaweb structure, as it is required to
accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas, Park, and
Drake (2017) suggested that most interactions (links) in ecological
networks are cryptic, \emph{i.e.} uncommon or hard to observe. This
argument echoes Jordano (2016): sampling ecological interactions is
difficult because it requires first the joint observation of two
species, and then the observation of their interaction. In addition, it
is generally expected that weak or rare interactions will be more
prevalent in networks than common or strong interactions (Csermely
2004); this is notably the case in food chains, wherein many weaker
interactions are key to the stability of a system (Neutel, Heesterbeek,
and de Ruiter 2002). In the light of these observations, we expect to
see an over-representation of low-probability (hereafter rare)
interactions under a model that accurately predicts interaction
probabilities.

Yet, the original metaweb definition, and indeed most past uses of
metawebs, was based on the presence/absence of interactions. Moving
towards \emph{probabilistic} metawebs, by representing interactions as
Bernoulli events (see \emph{e.g.} Poisot et al. 2016), offers the
opportunity to weigh these rare interactions appropriately. The inherent
plasticity of interactions is important to capture: there have been
documented instances of food webs undergoing rapid collapse/recovery
cycles over short periods of time (\emph{e.g.} Pedersen et al. 2017).
Furthermore, because the structure of the metaweb cannot be known in
advance, it is important to rely on predictive tools that do not assume
a specific network topology for link prediction (Gaucher, Klopp, and
Robin 2021), but are able to work on generalizations of the network that
capture statistical processes giving it its structure. These
considerations emphasize why metaweb predictions should focus on
quantitative (preferentially probabilistic) predictions, and this should
constrain the suite of models that are appropriate for prediction.
Binary classifiers based on probabilities have an extremely robust
methodology to validate them, and this applies naturally to the
prediction of interactions (Poisot 2023).

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al. 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between (Catchen et al. 2023). Although metawebs can
(and in practice likely do) include false positives, these are
statistically negligible compared to the false negatives. Furthermore,
Strydom et al. (2022) shows that t-SVD embedding is extremely robust to
(and able to detect) the presence of false positives. In a sense,
because most ecological interactions are elusive, we should consider the
direct consequences this has on sampling: once the common interactions
are documented, the effort required in documenting each rare interaction
will increase exponentially (Jordano 2016). Recent proposals in other
fields relying on machine learning approaches emphasize the idea that
algorithms meant to predict, through the assumption that they
approximate the process generating the data, can also act as data
generators (Hoffmann et al. 2019). High quality observational data can
be used to infer core rules underpinning network structure, and be
supplemented with synthetic data coming from predictive models trained
on them, thereby increasing the volume of information available for
analysis. Indeed, Strydom et al. (2021) suggested that knowing the
metaweb may render the prediction of local networks easier, because it
fixes an ``upper bound'' on which interactions can exist. In this
context, a probabilistic metaweb represents an aggregation of
informative priors on the biological feasibility of interactions, which
is usually hard to obtain yet has possibly the most potential to boost
our predictive ability of local networks (Bartomeus 2013; Bartomeus et
al. 2016). This would represent a departure from simple rules expressed
at the network scale (\emph{e.g.} Williams and Martinez 2000) to a view
of network prediction based on learning the rules that underpin
interactions \emph{and} their variability (Anubhav Gupta, Furrer, and
Petchey 2022).

@@

\caption{\label{fig-embedding}The embedding process (\textbf{A}) can
help to identify links (interactions) that may have been missed within
the original community (represented by the orange dashed arrows,
\textbf{B}). Transfer learning (\textbf{D}) allows for the prediction of
links (interactions) even when novel species (\textbf{C}) are included
alongside the original community. This is achieved with other
ecologically relevant predictors (\emph{e.g.} traits) in conjunction
with the known interactions to infer latent values (\textbf{E}).
Ultimately this allows us to predict links (interactions) for species
external from the original sample (blue dashed arrows) as well as
missing within sample links (\textbf{F}). Within this context the
predicted (and original) networks as well as the ecological predictors
used (green boxes) are products that can be quantified through
measurements in the field, whereas the embedded as well as imputed
matrices (purple box) are representative of a decomposition of the
interaction matrices onto the embedding space}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Box 1 - Graph Neural Networks}, bottomtitle=1mm, colback=white, breakable, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, left=2mm, arc=.35mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, leftrule=.75mm, rightrule=.15mm, titlerule=0mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6]

@@

different methods on a problem-specific basis, the fact that many

@@

general method of SVD. Note further that t-SNE has been included because

@@

t-SNE & nodes & statistical divergence & Hinton and Roweis (2002) &

@@

In Figure~\ref{fig-illustration-1}, we focus on some statistical checks
of the embedding. In panel \textbf{A}, we show that the averaged \(L_2\)
loss (\emph{i.e.} the mean of squared errors) between the empirical and
reconstructed metaweb decreases when the number of dimensions (rank) of
the subspace increases, with an inflection at 39 dimensions (out of 120
initially) according to the finite differences method. As discussed by
Runghen, Stouffer, and Dalla Riva (2021), there is often a trade-off
between the number of dimensions to use (more dimensions are more
computationally demanding) and the quality of the representation. In
panel \textbf{B}, we show the increase in cumulative variance explained
at each rank, and visualize that using 39 ranks explains about 70\% of
the variance in the empirical metaweb. This provides different
information from the \(L_2\) loss (which is averaged across
interactions), as it works on the eigenvalues of the embedding, and
therefore captures higher-level features of the network. In panel
\textbf{C}, we show positions of hosts and parasites on the first two
dimensions of the left and right subspaces. Note that these values
largely skew negative, because the first dimensions capture the coarse
structure of the network: most pairs of species do not interact, and
therefore have negative values. Finally in panel \textbf{D}, we show the
predicted weight (\emph{i.e.} the result of the multiplication of the
RDGP subspaces at a rank of 39) as a function of whether the
interactions are observed, not-observed, or unknown due to lack of
co-occurrence in the original dataset. This reveals that the observed
interactions have higher predicted weights, although there is some
overlap; the usual approach to identify potential interactions based on
this information would be a thresholding analysis, which is outside the
scope of this manuscript (and is done in the papers cited in this
illustration). Because the values returned from RDPG are not bound to
the unit interval, we performed a clamping of the weights to the unit
space, showing a one-inflation in documented interactions, and a
zero-inflation in other species pairs. Panel \textbf{D} specifically
shows that species pairs with no documented co-occurrence have weights
that are not distinguishable from species pairs with no documented
interactions; in other words, looking at the embedding, species that do
not co-occur are not easily distinguished from species that do not
interact. This suggests that (as befits a host-parasite model) the
ability to interact is a strong predictor of co-occurrence.

@@

In Figure~\ref{fig-illustration-2}, we relate the values of latent
variables for hosts to different ecologically-relevant data; we can
perform this additional step, because the results presented in
Figure~\ref{fig-illustration-1} show that we can extract an embedding of
the metaweb that captures enough variance to be relevant. Importantly,
this is true for both \(L_2\) loss (indicating that RDPG is able to
capture pairwise processes) and the cumulative variance explained
(indicating that RDPG is able to capture network-level structure), which
suggests that these approaches may allow to predict interactions
\emph{and} network structure. In panel \textbf{A}, we show that host
with a higher value on the first dimension have fewer parasites. This
relates to the body size of hosts in the \emph{PanTHERIA} database
(Jones et al. 2009), as shown in panel \textbf{B}: interestingly, the
position on the first axis is only weakly correlated to body mass of the
host; this matches well established results showing that body size/mass
is not always a direct predictor of parasite richness in terrestrial
mammals (Morand and Poulin 1998), a result we observe in panel
\textbf{C}. Finally, in panel \textbf{D}, we can see how different
taxonomic families occupy different positions on the first axis, with
\emph{e.g.} Sciuridae being biased towards higher values. These results
show how we can look for ecological informations in the output of
network embeddings, which can further be refined into the selection of
predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions
between species at a large spatial scale, typically a scale large enough
to be considered of biogeographic relevance (indeed, many of the
examples covered in the introduction span areas larger than a country,
some of them global). But as Herbert (1965) rightfully pointed out,
``{[}y{]}ou can't draw neat lines around planet-wide problems''; any
inference of a metaweb must therefore contend with several novel,
interwoven, families of problems. In this section, we outline three that
we think are particularly important, and discuss how they may be
addressed with subsequent data analysis or simulations, and how they
emerge in the specific context of using embeddings; some of these issues
are related to the application of these methods at the science-policy
interface. Adressing these considerations as part of the methodological
discussion is particularly important, as the construction of metawebs
can perpetuate legacies of biases in data (Box 2).

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic
point of view, the chances decrease of finding another area with enough
related species (through phylogenetic relatedness or similarity of
functional traits) to make a reliable inference. This is because
transfer requires similarity (Figure~\ref{fig-embedding}). A diagnostic
for the lack of similar species would likely be large confidence
intervals during estimation of the values in the low-rank space. In
other words, the representation of the original graph is difficult to
transfer to the new problem. Alternatively, if the initial metaweb is
too large (taxonomically), then the resulting embeddings would need to
represent interactions between taxonomic groups that are not present in
the new location. This would lead to a much higher variance in the
starting dataset, and to under-dispersion in the target dataset,
resulting in the potential under or over estimation of the strength of
new predicted interactions. Llewelyn et al. (2022) provided compelling
evidence for these situations by showing that, even at small spatial
scales, the transfer of information about interactions becomes more
challenging when areas rich with endemic species are considered. The
lack of well documented metawebs is currently preventing the development
of more concrete guidelines. The question of phylogenetic relatedness
and distribution is notably relevant if the metaweb is assembled in an
area with mostly endemic species (\emph{e.g.} a system that has
undergone recent radiation or that has remained in isolation for a long
period of time might not have an analogous system with which to draw
knowledge from), and as with every predictive algorithm, there is room
for the application of our best ecological judgement. Because this
problem relates to distribution of species in the geographic or
phylogenetic space, it can certainly be approached through assessing the
performance of embedding transfer in simulated starting/target species
pools.

@@

Predictive approaches in ecology, regardless of the scale at which they
are deployed and the intent of their deployment, originate in the
framework that contributed to the ongoing biodiversity crisis (Adam
2014) and reinforced environmental injustice (Choudry 2013; Domínguez
and Luoma 2020). The risk of embedding this legacy in our models is
real, especially when the impact of this legacy on species pools is
being increasingly documented. This problem can be addressed by
re-framing the way we interact with models, especially when models are
intended to support conservation actions. Particularly on territories
that were traditionally stewarded by Indigenous people, we must
interrogate how predictive approaches and the biases that underpin them
can be put to task in accompanying Indigenous principles of land
management (Eichhorn, Baker, and Griffiths 2019; No'kmaq et al. 2021).
The discussion of ``algorithm-in-the-loop'' approaches that is now
pervasive in the machine learning community provides examples of why
this is important. Human-algorithm interactions are notoriously
difficult and can yield adverse effects (Green and Chen 2019; Stevenson
and Doleac 2021), suggesting the need to systematically study them for
the specific purpose of, here, biodiversity governance. Improving the
algorithmic literacy of decision makers is part of the solution
(\emph{e.g.} Lamba et al. 2019; Mosebo Fernandes et al. 2020), as we can
reasonably expect that model outputs will be increasingly used to drive
policy decisions (Weiskopf et al. 2022). Our discussion of these
approaches need to go beyond the technical and statistical, and into the
governance consequences they can have. To embed data also embeds
historical and contemporary biases that acted on these data, both
because they shaped the ecological processes generating them (see Box
2), and the global processes leading to their measurement and
publication. For a domain as vast as species interaction networks, these
biases exist at multiple scales along the way, and a challenge for
prediction is not only to develop (or adopt) new quantitative tools, but
to assess the behavior of these tools in the proper context.

@@

Although promising, the application of embeddings to metaweb prediction
still involved several challenges. First, there is a need to understand
how to define a metaweb as a single, cohesive, unit of ecological
organisation. This is likely to have very different answers based on the
specific taxonomic group, temporal and spatial resolution, and question
being investigated. Second, there is a need to understand the scale at
which these predictions are relevant. Although we have documented many
cases of using embedding to fill gaps in the metaweb, these techniques
can likely be brought into a spatial (and possibly temporal) context.
The validation of these predictions will have to proceed jointly with
empirical sampling of interactions, but also with the design of
downsampling methods. Finally, there is a need for a greater
understanding of how biases in the data propagate to the predictions.
Because the volume of metawebs is currently low, and because graph
embeddings have not been commonly applied, we anticipate that this
discussion will take place organically in the coming years.
